Mahotas: Open source software for scriptable computer vision

### 2013-05-03 paper.tex
ENH call pyplot.show() at end of demo
ENH Read through. Several corrections

Mahotas is a computer vision library for the Python Programming Language
(versions 2.5 and up, including version 3 and up). It operates on numpy
arrays~\citep{numpystructure}. Therefore, it uses all the infrastructure built
by that project for storing information and performing basic manipulations and
computations. In particular, unlike libraries written in the C~Language or in
Java~\citep{Pietzsch15112012,Marcel:2010:TMP:1873951.1874254}, Mahotas does not need to define a
new image data structure, but uses the numpy array structure. Many basic
manipulation functionality that would otherwise be part of a computer vision
library are handled by numpy. For example, computing averages and other simple
statistics, handling multi-channel images, converting between types (integer
and floating point images are supported by mahotas) can all be performed with
numpy builtin functionality. For the user, this has the additional advantage
that they do not need to learn yet another set of functions.

It contains over 100~functions with functionality ranging from traditional
image filtering and morphological operations to more modern wavelet
decompositions and local feature computations. Additionally, by integrating
into the Python numeric ecosystem, users can use other packages in a seamless
way. In particular, mahotas does not implement any machine learning
functionality, but rather advises the user to use another, specialised package,
such as scikits-learn or milk.

@@

Numpy arrays contain data of a specific type, such \code{unsigned 8 bit
integer} or floating point numbers. While natural colour images are typically
8~bits, scientific data is often larger and processing can result in floating
point images. Mahotas works on all datatypes. This is performed without any
extra memory copies. Mahotas is heavily optimised for both speed and memory
usage (it can be used with very large arrays).

@@

We start by importing the necessary packages, including numpy and mahotas. We
also use \textit{scipy.cluster}, to demonstrate how the mahotas output can integrate
with a machine learning package.

@@

Finally, we can show the points in different colours.
\begin{python}
colors = np.array([
    [ 255,  25,   1],
    [203,  77,  37],
    [151, 129,  56],
    [ 99, 181,  52],
    [ 47, 233,   5]])
f2 = surf.show_surf(f, spoints[:64], cids, colors)
\end{python}

@@

\begin{python}
from matplotlib import pyplot as plt
plt.subplot(1,2,1)
plt.imshow(f)
plt.subplot(1,2,2)
plt.imshow(f2)
\end{python}

The easy interaction with matplotlib is another way in which we benefit from
the numpy-based ecosystem as mahotas does not need to support interacting with
a graphical system to display images.

@@

So, for example, this is how \emph{erode} is implemented.\footnote{This is the
generic version of the code. Erode, like a few other functions, has two
versions, a fast version, limited to two-dimensional, contiguous, images and a
generic one presented here. The selection between the two implementations is
done automatically for the user.} \code{py\_erode} consists mostly of
boiler-plate code:

@@

\begin{description}
\item[\code{gil\_release}] This is a ``resource-acquisition is object
initialisation'' (\textsc{raii})\footnote{\textsc{Raii} is a design pattern in
\cpp{}, or other languages with scope linked deterministic object destruction,
such as D, where a resource is represented by an object, whose constructor
acquires it and whose destructor releases it. This guarantees that the object
is correctly released even if the scope is left through an exception
\citep{Stroustrup1994}.} object that releases the Python global interpreter lock
(\textsc{gil})\footnote{In the CPython interpreter, the most commonly used
implementation of Python, there is a global lock for many Python related
functionality, which limits parallelism.} in its constructor and gets it back
in its destructor. Normally, the template function will release the
\textsc{gil} after the Python-specific code is done. This allows several
mahotas functions to run concurrently.
\item[\code{array}] This is a thin wrapper around \code{PyArrayObject}, the raw
numpy data type, which has iterators which resemble the \cpp{} standard
library. It also handles type-casting internally, making the code type-safer.
This is also a \textsc{raii} object in terms of managing Python reference
counts. In mahotas debug builds, this object additionally adds several checks
to all the memory acesses.
\item[\code{filter\_iterator}] This was adapted from code in the
\code{scipy.ndimage} packages and it is useful to iterate over an image and use
a centered filter around each pixel (it keeps track of all of the boundary
conditions).
\end{description}

@@

Table~\ref{tab:efficiency} shows timings for different operations. These were
normalized to multiples of the time it takes to go over the image and find its
maximum pixel value (this was done using \code{numpy.max(image)}). The
measurements shown were obtained on an Intel 64~bit system, running Ubuntu
Linux. However, due to the normalization, measurements obtained on another
system (Intel 32~bits running Mac OS) were qualitatively similar.

@@

OpenCV is the fastest library, but this comes at the cost of some flexibility.
Arguments to its functions must be of the exact expected type and it is
possible to crash the interpreter if types do match the expected type (in the
other libraries, including mahotas, all types are checked and an exception is
generated which can be caught by user code). This is particularly relevant for
interactive use as the user is often exploring and is willing to pay the speed
cost of a few extra type checks and conversions to avoid a hard-crash.

@@

Mahotas includes a complete automated suite of unit tests, which tests all
functionality and include several regression tests. There are no known bugs in
version~1.0. In fact, no releases have ever been performed with known bugs.
Naturally, bugs were, occasionally, discovered in released versions, but
corrected before the next release.

@@

Mahotas is implemented in \cpp{}, as the standard Python interpreter is too
slow for a direct Python implementation. However, all of the Python interface
code is hand-written, as opposed to using automatic interface generators like
Swig~\citep{Beazley2003599}. This is more work initially, but the end result is
of much higher quality, especially when it comes to giving useful error
messages (e.g., when a type mismatch occurs, an automatic system will often be
forced to resort to a generic message as it does not have any knowledge of what
the arguments mean besides their automatically inferred types).

@@

Mahotas includes code ported and incorporated from other projects. Initially,
it was used in reproducing the functionality in the Subcellular Location Image
Classifier (SLIC) tool from Robert F.\ Murphy's
Lab~\citep{springerlink:10.1007/978-0-387-45524-2_47} and the initial versions
of mahotas were designed explicitly to support that functionality. The
\textsc{surf} implementation is a port from the code from
\textit{dlib},\footnote{Dlib's webpage is at \url{http://dlib.net}.} a very
good \cpp{} library by Davis King. I also gleaned some insight into the
implementation of these features from Christopher Evan's OpenSURF library and
its documentation \citep{evans2009}.\footnote{OpenSURF is available at
\url{http://www.chrisevansdev.com/computer-vision-opensurf.html}, where several
documents describe details of the implementation.} The code which interfaces
with the FreeImage library, was written by Zachary Pincus and some of the
support code was written by Peter J. Verveer for the \code{scipy.ndimage}
project. All of these contributions were integrated while respecting the
software licenses under which the original code had been released. Robert Webb,
a summer student at Carnegie Mellon University, worked with me on the initial
local binary patterns implementation. Finally, I thank the several users who
have reported bugs, submitted fixes, and participated on the project
mailing list.

@@

\textbf{Funding}: I was supported in my work by the Funda\c c\~{a}o para a
Ci\^encia e Tecnologia (grants SFRH/BD/37535/2007 and
PTDC/SAU-GMG/115652/2008), by NIH grant GM078622 (to Robert F.\ Murphy), by a
grant from the Scaife Foundation, by the HHMI Interfaces Initiative, and by a
grant from the Siebel Scholars Foundation.

>>>>

Mahotas is a computer vision library for the Python Programming Language
(versions 2.5 and up, including version 3 and up). It operates on numpy
arrays~\citep{numpystructure}. Therefore, it uses all the infrastructure built
by that project for storing information and performing basic manipulations and
computations. In particular, unlike libraries written in the C~Language or in
Java~\citep{Pietzsch15112012,Marcel:2010:TMP:1873951.1874254}, Mahotas does not need to define a
new image data structure, but uses the numpy array structure. Many basic
manipulation functionality that would otherwise be part of a computer vision
library are handled by numpy. For example, computing averages and other simple
statistics, handling multi-channel images, converting between types (integer
and floating point images are supported by mahotas) can all be performed with
numpy builtin functionality. For the users, this has the additional advantage
that they do not need to learn new interfaces.

Mahotas contains over 100~functions with functionality ranging from traditional
image filtering and morphological operations to more modern wavelet
decompositions and local feature computations. Additionally, by integrating
into the Python numeric ecosystem, users can use other packages in a seamless
way. In particular, mahotas does not implement any machine learning
functionality, but rather advises the user to use another, specialised package,
such as scikits-learn or milk.

@@

Numpy arrays contain data of a specific type, such \code{unsigned 8 bit
integer} or floating point numbers. While natural colour images are typically
8~bits, scientific data is often larger (12 and 16~bit formats are common).
Processing can generate floating point images, for example, subtracting the
mean value to every pixel; the mean will be a floating point number even if the
original image was integral. Mahotas works on all datatypes. This is performed
without any extra memory copies. Mahotas is heavily optimised for both speed
and memory usage (it can be used with very large arrays).

@@

We start by importing the necessary packages, including numpy and mahotas. We
also use \code{scipy.cluster}, to demonstrate how the mahotas output can
integrate with a machine learning package.

@@

Finally, we can show the points in different colours. In order to avoid a very
cluttered image, we will only plot the first 64~regions.

\begin{python}
colors = np.array([
    [ 255,  25,   1],
    [203,  77,  37],
    [151, 129,  56],
    [ 99, 181,  52],
    [ 47, 233,   5]])
f2 = surf.show_surf(f, spoints[:64], cids, colors)
\end{python}

@@

\begin{python}
from matplotlib import pyplot as plt
plt.subplot(1,2,1)
plt.imshow(f)
plt.subplot(1,2,2)
plt.imshow(f2)
plt.show()
\end{python}

The easy interaction with matplotlib is another way in which we benefit from
the numpy-based ecosystem. Mahotas does not need to support interacting with a
graphical system to display images.

@@

So, for example, this is how \emph{erode} is implemented.\footnote{This is the
generic version of the code. Erode, like a few other functions, has two
versions, a fast version, limited to two-dimensional, contiguous, images and a
generic one presented here. The selection between the two implementations is
done automatically.} \code{py\_erode} consists mostly of
boiler-plate code:

@@

\begin{description}
\item[\code{gil\_release}] This is a ``resource-acquisition is object
initialisation'' (\textsc{raii})\footnote{\textsc{Raii} is a design pattern in
\cpp{}, or other languages with scope linked deterministic object destruction,
such as D, where a resource is represented by an object, whose constructor
acquires it and whose destructor releases it. This guarantees that the object
is correctly released even if the scope is left through an exception
\citep{Stroustrup1994}.} object that releases the Python global interpreter lock
(\textsc{gil})\footnote{In the CPython interpreter, the most commonly used
implementation of Python, there is a global lock for many Python related
functionality, which limits parallelism. By releasing the \textsc{gil}, mahotas
allows another thread to run concurrently.} in its constructor and gets it back
in its destructor. Normally, the template function will release the
\textsc{gil} after the Python-specific code is done. This allows several
mahotas functions to run concurrently.
\item[\code{array}] This is a thin wrapper around \code{PyArrayObject}, the raw
numpy data type, which has iterators which resemble the \cpp{} standard
library. It also handles type-casting internally, making the code type-safer.
This is also a \textsc{raii} object in terms of managing Python reference
counts. In mahotas debug builds, this object additionally adds several checks
to all the memory acesses.
\item[\code{filter\_iterator}] This was adapted from code in the
\code{scipy.ndimage} packages and it is useful to iterate over an image and use
a centered filter around each pixel (it keeps track of all of the boundary
conditions).
\end{description}

@@

Table~\ref{tab:efficiency} shows timings for different operations. These were
normalized to multiples of the time it takes to go over the image and find its
maximum pixel value (using the expression \code{numpy.max(image)}). The
measurements shown were obtained on an Intel 64~bit system, running Ubuntu
Linux. Due to the normalization, measurements obtained on another system (Intel
32~bits running Mac OS) were qualitatively similar.

@@

OpenCV is the fastest library, but this comes at the cost of some flexibility.
Arguments to its functions must be of the exact expected type and it is
possible to crash the interpreter if types do match the expected type (in the
other libraries, including mahotas, all types are checked and an exception is
generated which can be caught by user code). This is particularly relevant for
interactive use as the user is often exploring and is willing to pay the speed
cost of a few extra type checks to avoid a hard-crash.

@@

Mahotas includes a complete automated suite of unit tests, which tests all
functionality and include several regression tests. There are no known bugs in
version~1.0. In fact, no releases have every contained any known bugs.
Naturally, bugs were, occasionally, discovered in released versions, but
corrected before the next release.

@@

Mahotas is implemented in \cpp{}, as the standard Python interpreter is too
slow for a direct Python implementation. However, all of the Python interface
code is hand-written, as opposed to using automatic interface generators like
Swig~\citep{Beazley2003599}. This is more work initially, but the end result is
of much higher quality, especially when it comes to giving useful error
messages. When a type mismatch occurs, an automatic system will often be forced
to resort to a generic message as it does not have any knowledge of what the
arguments mean to the user. It will only know their automatically inferred
types. A hand written system can also automatically convert arguments when
meaningful and be more flexible without completely losing type checking.

@@

Mahotas includes code ported and incorporated from other projects. Initially,
it was used in reproducing the functionality in the Subcellular Location Image
Classifier (SLIC) tool from Robert F.\ Murphy's
Lab~\citep{springerlink:10.1007/978-0-387-45524-2_47} and the initial versions
of mahotas were designed explicitly to support that functionality. The
\textsc{surf} implementation is a port from the code from
\textit{dlib},\footnote{Dlib's webpage is at \url{http://dlib.net}.} an
excellent \cpp{} library by Davis King. I also gleaned some insight into the
implementation of these features from Christopher Evan's OpenSURF library and
its documentation \citep{evans2009}.\footnote{OpenSURF is available at
\url{http://www.chrisevansdev.com/computer-vision-opensurf.html}, where several
documents describe details of the implementation.} The code which interfaces
with the FreeImage library, was written by Zachary Pincus and some of the
support code was written by Peter J. Verveer for the \code{scipy.ndimage}
project. All of these contributions were integrated while respecting the
software licenses under which the original code had been released. Robert Webb,
a summer student at Carnegie Mellon University, worked with me on the initial
local binary patterns implementation. Finally, I thank the several users who
have reported bugs, submitted fixes, and participated on the project
mailing list.

@@

\textbf{Funding}: I was supported in my work by the Funda\c c\~{a}o para a
Ci\^encia e Tecnologia (grant SFRH/BD/37535/2007 to himself and grant
PTDC/SAU-GMG/115652/2009 to Musa Mhlanga), by NIH grant GM078622 (to Robert F.\
Murphy), by a grant from the Scaife Foundation, by the HHMI Interfaces
Initiative, and by a grant from the Siebel Scholars Foundation.
